Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-14 writeup/conStruct.tex
edited up to empirical apps

Another way to describe how additional layers contribute to model fit
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers, 
so that they will have a low contribution to overall covariance.
We use the connection of allelic covariance to the probability of agreement (identity) between genotypes
to calculate a measure of overall contribution of each layer to the model.
XXX PLR XXX
To calculate the overall contribution of the $k$th layer to the model,
we use the connection between allelic covariance and

The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The contribution of the $k$th layer is its absolute contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this contribution, see Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

\subsection*{Simulations}
We first test the method with scenarios generated using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
To generate discrete population structure, we simulate $K$ distinct populations,
each of which split from a common ancestor $\tau_{\text{s}}$ units of coalescent time in the past,
without subsequent migration between them.
To generate continuous differentiation within each layer,
at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonizes an
independent lattice of demes,
for which we use a stepping stone model with symmetric migration 
to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past 
we generate a single dataset 
by collapsing those $K$ discrete layers into a single grid of demes 
that have various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapse the layers together using random admixture proportions 
that are spatially autocorrelated (see Appendix \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers,
in each sampling 10,000 unlinked loci from each of 20 haploid individuals from every deme.

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the analyses of the dataset simulated under $K=1$, 
contrasting the results of the nonspatial model runs with 2 - 4 layers 
with those of spatial model with 2 - 4 layers.
In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
To confirm that results from our nonspatial model were qualitatively similar to those of 
other nonspatial model-based clustering algorithms, 
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}).

A number of results are immediately clear from these tests with simulated data.
First, \texttt{conStruct} performs well on all simulated scenarios.
Across all values of K and for all simulations, 
the spatial model is correctly preferred over the nonspatial model
(Fig \ref{sim_xvals}). Second, the predictive accuracy of the nonspatial model increases 
with the number of layers in the model, 
as it describes continuous, spatial patterns of relatedness between demes within the same cluster 
using gradients of admixture across more and more discrete clusters.
This behavior of the nonspatial model is highlighted in 
the analyses of data simulated under $K=1$ (Fig \ref{simK1_pies} and \ref{simK1_nsp_pies}):
as $K$ in the model increases, 
the sampled area is increasingly divvied up between clusters 
with ``centers of mass" in the $K$ geographical extremes of the simulated space.
At $K=2$, the clusters are centered in opposite corners on the map; 
at $K=3$, in the lower two corners and the top half; 
at $K=4$, all four corners, and so on.
These results are qualitatively similar to those generated using fastSTRUCTURE \citep{fastStructure}, 
which also describes continuous differentiation using admixture between clusters
 (Fig \ref{fastStr_simK1}-\ref{fastStr_simK3}).
While it is true that each non-spatial cluster is genetically more similar within itself
than it is to other clusters,
we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
either plateaus at or declines from the true value of $K$ used to simulate the data 
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).
And, as $K$ in the spatial model increases, 
deme membership in layers beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

Finally, we calculated layer contribution for 
analyses run in each simulation 
(Fig \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}).
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional clusters in the nonspatial analyses continue to
be important for describing variation, 
as they are forced to partition up the spatial patterns within clusters.

>>>>

Another way to describe the practical significance of additional layers
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers if they are present, 
so that they will have a low contribution to overall covariance.
To do this, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some background probability,
but this probability of agreement is increased on any segment of genome 
that both have inherited through the same layer.
(The amount it increases depends on how far apart they are.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

\subsection*{Simulations}

To test the method, we first generated data using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture, spatially autocorrelated proportions 
(see Appendix \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers,
in each sampling 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation to compare the predictive performance.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, using $K=2--4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

\plr{rearranged this to focus more on take-homes}

% In Fig \ref{simK1_pies}, we show the results of the dataset simulated with $K=1$, 
% contrasting the results of the nonspatial model runs with 2--4 layers 
% with those of spatial model with 2--4 layers.
% 
% In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
% The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
% Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
% 
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
% we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

With these simulations,
spatial \texttt{conStruct} does not create spurious clusters when there are none:
Figures \ref{simK1_pies}, \ref{simK2_sp_pies}, and \ref{simK3_sp_pies} 
show that subsequent layers beyond the number used for simulation are unused 
when data is simulated with $K=1$ but analyzed with $K>1$ --
the layers are present, but contribute very little to any population.
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

In contrast, the nonspatial model describes geographic variation
using gradients of admixture between more and more discrete clusters
to better approximate the continuous, spatial patterns of relatedness
(depicted in Figs \ref{simK1_pies} and \ref{simK1_nsp_pies}).
% as $K$ in the model increases, 
% the sampled area is increasingly divvied up between clusters 
% with ``centers of mass" in the $K$ geographical extremes of the simulated space.
% At $K=2$, the clusters are centered in opposite corners on the map; 
% at $K=3$, in the lower two corners and the top half; 
% at $K=4$, all four corners, and so on.
The fastSTRUCTURE results are qualitatively similar, as shown in Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}.
Each non-spatial cluster is in truth genetically more similar within itself
than it is to other clusters,
but we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}),
while the predictive accuracy of the nonspatial model increases as subsequent clusters are added.

% In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
% either plateaus at or declines from the true value of $K$ used to simulate the data 
% (Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).

The unimportance of spurious layers can be seen in plots of layer contributions,
Figs \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}.
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional clusters in the nonspatial analyses continue to contribute substantially.
